%% ============================================================================
%%
%%  Master's thesis
%%
%%  Author: FORNAVN EFTERNAVN
%%
%%  Chapter 2: Another topic....
%% ============================================================================

\chapter{BBN code}
\label{chap:BBNcode}

\section{History of BBN codes}
\label{sec:BBN_history}

The concept of Big Bang nucleosynthesis is almost as old as the Big bang theory itself, with the with it first being proposed in the paper by \cite{Gamov48}. However interest waned, as stellar nucleosynthesis proved itself much more useful in explaining present element abundances. 

With the discovery of the CMB in 1965, new attention was brought to the early universe. Only a year later Peebles showed how simple BBN physics could be used to explain the high helium abundance, unaccounted for by stellar nucreosynthesis \cite{Peebles66}. Soon after Wagoner created and refined the first proper bbn code, described in a series of classic papers \cite{Wagoner67} \cite{Wagoner69} \cite{Wagoner72}.


By the late 80's the wagoner code was severly outdated. With multiple ineffeciencies due to among other things, the fact that it was originally designed to run on punch cards. This inspired Lawrence Kawano to create the now ubiquitous NUC123, colloquially know as the Kawano code. Which set the gold Standard for all future BBN codes. 



\subsection{Wagoner}
\label{sec:Wagoner}

\subsection{Kawano}
\label{sec:Kawano}

\subsection{Modern codes}
\label{sec:Modern_codes}
PArthENoPE, AlterBBN, PRIMAT


\subsection{AlterBBN}
\label{sec:AlterBBN}

AlterBBN is written in c and based on Kawano's NUC123. It maintains the same basic structure and integration method, though it uses natural units for everything but the reaction network. However they defines energy in GeV rather than MeV. What seperates AlterBBN for other codes is that as the name implies, it allows the use of alternate cosmological models and parametes. Therefore this code is especially well suited for testing the effects these alterrations have on final abundances.